Historical aerial surveys map long-term changes of forest cover and structure in the central Congo Basin

### 2020-01-23 manuscript/main.tex
revision

% You need to save the "mdpi.cls" and "mdpi.bst" files into the same folder as this template file.

%=================================================================
\documentclass[remote sensing,article,submit,moreauthors,pdftex,10pt,a4paper]{mdpi}

@@

%
%--------------------
% Class Options:
%--------------------
% journal
%----------
% Choose between the following MDPI journals:
% actuators, addictions, admsci, aerospace, agriculture, agronomy, algorithms, animals, antibiotics, antibodies, antioxidants, applsci, arts, asi, atmosphere, atoms, axioms, batteries, bdcc, behavsci, beverages, bioengineering, biology, biomedicines, biomimetics, biomolecules, biosensors, brainsci, buildings, carbon, cancers, catalysts, cells, ceramics, challenges, chemengineering, chemosensors, children, chromatography, climate, coatings, colloids, computation, computers, condensedmatter, cosmetics, cryptography, crystals, cybersecurity, data, dentistry, designs, diagnostics, diseases, diversity, econometrics, economies, education, electrochemistry, electronics, energies, entropy, environments, epigenomes, est, fermentation, fibers, fire, fishes, fluids, foods, forests, fractalfract, futureinternet, galaxies, games, gastrointestdisord, gels, genealogy, genes, geosciences, geriatrics, hazardousmatters, healthcare, heritage, highthroughput, horticulturae, humanities, hydrology, informatics, information, infrastructures, inorganics, insects, instruments, ijerph, ijfs, ijms, ijgi, ijtpp, inventions, j, jcdd, jcm, jcs, jdb, jfb, jfmk, jimaging, jof, jintelligence, jlpea, jmmp, jmse, jpm, jrfm, jsan, land, languages, laws, life, literature, logistics, lubricants, machines, magnetochemistry, make, marinedrugs, materials, mathematics, mca, mti, medsci, medicines, membranes, metabolites, metals, microarrays, micromachines, microorganisms, minerals, modelling, molbank, molecules, mps, nanomaterials, ncrna, neonatalscreening, neuroglia, nitrogen, nutrients, ohbm, particles, pathogens, pharmaceuticals, pharmaceutics, pharmacy, philosophies, photonics, plants, plasma, polymers, polysaccharides, preprints, proceedings, processes, proteomes, publications, quaternary, qubs, recycling, religions, remotesensing, reports, resources, risks, robotics, safety, scipharm, sensors, separations, sexes, sinusitis, socsci, societies, soils, sports, standards, stats, surgeries, sustainability, symmetry, systems, technologies, toxics, toxins, tropicalmed, universe, urbansci, vaccines, vetsci, vibration, viruses, vision, water, wem
%---------
% article
%---------
% The default type of manuscript is article, but can be replaced by:
% abstract, addendum, article, benchmark, book, bookreview, briefreport, casereport, changes, comment, commentary, communication, conceptpaper, correction, conferenceproceedings, conferencereport, expressionofconcern, meetingreport, creative, datadescriptor, discussion, editorial, essay, erratum, hypothesis, interestingimages, letter, meetingreport, newbookreceived, opinion, obituary, projectreport, reply, reprint, retraction, review, perspective, protocol, shortnote, supfile, technicalnote, viewpoint
% supfile = supplementary materials
% protocol: If you are preparing a "Protocol" paper, please refer to http://www.mdpi.com/journal/mps/instructions for details on its expected structure and content.
%----------
% submit
%----------
% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted. This will only make changes to the frontpage (e.g. the logo of the journal will get visible), the headings, and the copyright information. Also, line numbering will be removed. Journal info and pagination for accepted papers will also be assigned by the Editorial Office.
%------------------
% moreauthors
%------------------
% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.
%---------
% pdftex
%---------
% The option pdftex is for use with pdfLaTeX. If eps figures are used, remove the option pdftex and use LaTeX and dvi2pdf.

%=================================================================
\firstpage{1}
\makeatletter
\setcounter{page}{\@firstpage}
\makeatother
\articlenumber{x}
\doinum{10.3390/------}
\pubvolume{xx}
\pubyear{2018}
\copyrightyear{2018}
\externaleditor{Academic Editor: name}
\history{Received: date; Accepted: date; Published: date}

@@

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{Land Use and Land Cover change (LULCC) of African rainforest contribute
to global carbon emissions. Yet, most historical estimates of LULCC and
their carbon emissions rely on non-spatially explicity data in the
pre-satellite era (\textless{} 1972). Past studies have expanded this
pre-satellite time frame with declassified satellite surveillance data
or aerial surveys, but none cover the Congo Basin. Here, we use
historical aerial survey photos to map the extent and structure of LULCC
around Yangambi, central Congo Basin, in 1958. Our study leveraged
Structure-from-Motion to generate a large orthomosaic covering 828
km\textsuperscript{2}, counting 733 million pixels, at a sub meter
resolution and geo-referenced to \textasciitilde{}4.7 \(\pm\) 4.3 m.
Primary forest in the orthomosaic was classified with a Convolutional
Neural Network Deep Learning approach. We used these data to quantify
LULCC, landscape and canopy complexity and scale above ground biomass
between historical and contemporary condition. We show a shift from
historical highly structured industrial deforestation in 1958 (162
km\textsuperscript{2}), to contemporary smallholder farming clearing (88
km\textsuperscript{2}), increasing landscape fragmentation and forest
edges. Efforts to quantify canopy texture and their link to carbon
storage had limited to no success. Our analysis provides an insight in
key spatial and temporal patterns of deforestation and reforestation at
a multi-decadal scale and provide a historical context to land-cover and
land-use change spatial patterns for past and ongoing field research in
the area.}

@@

@@

\usepackage{wrapfig} \usepackage{graphics} \usepackage{setspace}

\doublespacing

\usepackage{gensymb} \usepackage{booktabs}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Only for the journal Gels: Please place the Experimental Section after the Conclusions

@@

The functioning of a tropical forest, and its capacity to sequester
carbon, is heavily influenced by anthropogenic disturbances such as
selective logging, clear felling, or slash and burn practices, all
increasing forest fragmentation. This increased pace in the loss of
rainforest cover \citep{butsic2015} is predicted to negatively impact
the function of these tropical forests, and their future capacity to
sequester carbon, as anthropogenic disturbances such as selective
logging, clear felling, or slash and burn practices collectively
contribute to increasing forest fragmentation.

These forest fragments and their edges contain fewer large trees
\citep{laurance2000}, in part due to their hotter and brighter
microclimates \citep{magnago2015}. Similarly, species composition and
biodiversity \citep{poorter2006} are negatively affected. Spatial
disturbances also have an explicit temporal component, where the time
since disturbance affects the strength of the effects
\citep{didham1999}. Deforestation and gap formation also influence the
overall canopy structure, where pioneer species grow taller and with
narrower crowns \citep{poorter2006, peres2006} affecting the forest
light regime and carbon dynamics \citep{stark2012}. As such,
anthropogenic factors through land-use and land cover change (LULCC)
heavily affect forest structure and functioning
\citep{fauset2017, brinck2017, didham1999}. In the tropics the majority
of emissions originate from deforestation and forest degradation
\citep{hansen2013, houghton2012, tyukavina2015} contributing to 10-15\%
of the total global emissions \citep{vanderwerf2009} and concomitant
losses in biodiversity losses \citep{barlow2016, vandeperre2018} and how
LULCC is evolving over time. Although the African rainforest is the
second largest on Earth and covers \textasciitilde{}630 million ha,
representing up to 66 Pg of carbon storage \citep{Lewis2009b}, and
currently loses forest at an increasing pace \citep{butsic2015}. Yet,
the tropics is now becoming a large source of CO\_2 to the atmosphere as
deforestation and forest degradation
\citep{hansen2013, houghton2012, tyukavina2015} contribute up to 10-15\%
of the total global CO2 emissions \citep{vanderwerf2009}. Furthermore,
recent estimates show that 31\% of recent carbon emissions are caused by
edge effects alone \citep{brinck2017}. In this context, there are very
few long term forest inventory plots that presently exist in the central
Congo Basin \citep{Verbeeck2011a, lewis2013}. This further limits the
accuracy of current biomass estimates and challenges our understanding
of how forest structure and function changes in the short and long term
in these ecosystems \citep{Kearsley2013, Doetterl2015}.

Historical estimates of LULCC, and their carbon emissions, exist
\citep{houghton2012, ramankutty1999, defries2002, sader1988}, however
they generally rely on non-spatially explicit data in the pre-satellite
era (\textless{} 1972) \citep{houghton2010}. These analysis often use
FAO statistics, historical land-use sources, and/or population and
socio-economic data \citep{houghton2012}, as limited earlier spatial
data sources are available \citep{hansen2013, achard2014}. For example,
the \citet{houghton2010} long-term study used FAO and national censuses
alongside a bookkeeping model to estimate carbon emissions from LULCC.
Other studies use remote sensing data, but are limited in time to recent
decades \citep{defries2002, houghton2012}. These proxy based studies
have a limited view on long-term direct and indirect spatial effects of
deforestation. As such, both the spatial structure and the fragmentation
pattern of a forest combined with its life history has a profound effect
on carbon emssisions. However, in this context remote sensing
assessments provide more accurate spatial information to help estimate
carbon emissions \citep{mitchard2018}.

Past studies expanded knowledge during the pre-satellite time frame
through the use of now declassified satellite surveillance data to map
deforestation in the US, central Brazil, and post world war II Germany
\citep{song2015, nita2018}. In some cases these records could extend
back in time even further using historical aerial survey images. In
northern Europe especially these data are well known and used, for
example to map the extent of forest cover across \textasciitilde{}100
km\textsuperscript{2} regularly, and at times using automated workflows
\citep{giordano2018}. Survey data across the African continent is less
common or inaccessible or both. Some studies do exist, both
\citet{buitenwerf2012} and \citet{okeke2006} used aerial survey images
to map vegetation changes in South African savannas and the Adulam
Nature Reserve, respectively. Whilst \citet{frankl2015} used 20 aerial
survey images to map the Ethiopian highlands of 1935. High resolution
aerial or surveillance satellite images provide scientists tools to
monitor forest extent and structure over long time frames. Moreover,
canopy texture from aerial survey images has been successfully linked to
above ground biomass \citep{Ploton2012, Couteron2005, Barbier2010}. Thus
using texture based metrics obtained from (high resolution) remote
sensing data opens up possibilities to link historical aerial footage to
carbon density across large spatial extents \citep{Couteron2005}.

Given the impact of LULCC on the structure and functioning of tropical
forests, and their influence on both carbon dynamics
\citep{Kearsley2013} and biodiversity \citep{vandeperre2018}, accurate
long term reporting of historical forest cover in the pre-satellite era
is an imperative \citep{willcock2016}. In this study we use a
combination of historical aerial photography and contemporary remote
sensing data to map long-term changes in the extent and structure of the
tropical forest surrounding Yangambi, in the central Congo Basin at the
beginning of 1958 and the start of the antrophocene \citep{lewis2015}.
Our analysis provides a historical insight into important LULCC spatial
patterns in Yangambi, such as fragmentation and edge complexity. Our
analysis further contextualises the influence of changes in the forest's
life history on past and current research into Above Ground Carbon (AGC)
storage \citep{Kearsley2013} and biodiversity \citep{vandeperre2018} in
the central Congo Basin.

@@

Data for the central Congo Basin region, surrounding Kisangani, were
collected in several flights during the dry season of 1958 and 1959
(from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959
respectively, see Appendix Figure 1) to generate topographic maps of the
area, supervised by the ``Institut Géographique du Congo Belge'' in
Kinshasa (then Léopoldville). Black-and-White infrared images (0.4 - 0.9
\(\mu\)m) were gathered along flight paths running mostly from west to
east, between 9 - 11h local time. Along a flight path continuous images
were taken using a Wild Heerburgg RC5a (currently Leica Geosystems) with
an Aviogon lens assembly (114.83mm / f 5.6, with a 90\degree view angle)
resulting in square photo negative of 180 by 180 mm. Flights were flown
at an average absolute altitude of \textasciitilde{}5200 m above sea
level, covering roughly 18 530 km\textsuperscript{2} at an approximate
scale of 1/40 000. The use of the integrated autograph system ensured
timely acquisition of pictures with a precise overlap
(\textasciitilde{}1/3) between images. This large overlap between images
together with flight parameters would allow post-processing, using
stereographs, to create accurate topographic maps. Original data from
this campaign are stored in the
\href{https://www.africamuseum.be/en}{Royal Museum for Central Africa}
in Tervuren, Belgium (Figure 1).

\begin{figure}
\includegraphics[width=1\linewidth]{./figures/print_examples} \caption{A box of historical aerial photographs (left) and a single aerial photograph (right) showing part of the Congo river. Note the meta-data provided in the margins of the image such as acquisition time and fleight height and the solar glare on the Congo river.}\label{fig:unnamed-chunk-1}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{./figures/flight_paths} \caption{Overview of the historical flight paths during aerial photo acquisition and ancillary data used in this study. The bounding box of the orthomosaic data presented in this study is shown as a rectangle (23x36 km). The outline of a recent high-resolution Geo-eye panchromatic image is shown as a dashed dark grey rectangle (10x10km). The location of various permanent sampling plots are shown as x, +, and open squares and triangles for the mixed, mono-dominant and edge plots respectively. The grey polygon delineates the current day Yangambi Man and Biosphere reserve. The inset, top right, situates the greater Yangabmi region with the DR Congo. The full flight plan and details are shown in Appendix Figure 1 and 2}\label{fig:unnamed-chunk-2}
\end{figure}

@@

All selected images, covering the Yangambi area, were contact prints as
original negatives of the prints were not available. Images were scanned
at a resolution exceeding their original resolution (or grain) at the
maximal physical resolution of an Epson A3 flatbed scanner (i.e.~2400
dpi or 160MP per image) and saved as lossless tiff images. Data were
normalized using contrast limited histogram equalization
\citep{zuiderveld1994} with a window size of 32 and a clip limit of 1.5.
Fiduciary marks were used to rectify and downsample the images into
square 7700x7700 pixel images (\textasciitilde{}1200 dpi, 81MP). This
resulted in a dataset with digital images at a resolution that remained
above the visible grain of the photographs, whilst the reduced image
size facilitated easier file handling and processing speed.

@@

We calculated the orthomosaic using a low resolution point cloud and
digital elevation map (DEM). Additional ground control points were
provided to assist in the referencing of image and constrain the
optimization routine used in the SfM algorithm. Ground control points
consisted of rooftop edges of permanent structures which could be
verified in both old and new aerial imagery (i.e.~ESRI World Imagery).
Although clouds were removed during the SfM routine we did not mask all
clouds in the final orthomosaic to maximize forest coverage. The final
scene was cropped, to provide consistent wall to wall coverage of the
reconstructed scene. The orthomosaic was exported as a geotiff for
further georeferencing in QGIS \citep{qgisdevelopmentteam2018} using the
georeferencer plugin (version 3.1.9) and additional ESRI World Imagery
high resolution reference data. We used 3\textsuperscript{th} degree
polynomial and 16 ground control points to correct the final image.
Ground control points, raw image data and final processed image are
provided in addition to measures of uncertainty such as mean root mean
squared (RMSE), mean and median error across all ground control points.
All subsequent analysis are executed on the final geo-referenced
orthomosaic or subsets of it.

@@

\hypertarget{classifying-forest-cover}{%
\subsubsection{Classifying forest
cover}\label{classifying-forest-cover}}

We automatically delineated all natural forest in the historical data,
thus excluding tree plantations, thinned or deteriorated forest stands,
fields and buildings. We used the Unet Convolutional Neural Net (CNN,
\citet{ronneberger2015}) architecture implemented in Keras
\citep{chollet2015} with an efficientnetb3 pre-processing backbone
\citep{yakubovskiy2019} running on TensorFlow \citep{martinabadi2015} to
train a binary classifier (i.e.~forest or non-forested). This
methodology is increasingly being used to automate pixel-level
classification in (color) digital photography data \citep{buscombe2018}.
Training data were collected from the orthomosaic by randomly selecting
513 pixel square tiles from homogeneous forested or non-forested areas
within the historical orthomosaic. Homogeneous tiles were combined in
synthetic landscapes using a random gaussian field based binary mask
(Figure 3). We generated 5000 synthetic landscapes for training, while
500 landscapes were generated for both the validation and the testing
dataset. Source tiles did not repeat across datasets to limit
overfitting. In order to limit stitch line misclassifications, along the
seams of mosaicked images, we created synthetic landscapes with
different forest tiles to mimick forest texture transitions. We applied
this technique to 10\% of the generated synthetic landscapes. The CNN
model was trained for 100 epochs with a batch size of 30 using Adam
optimization \citep{kingma2017}, maximizing the Intersect-over-Union
(IoU) using Sørensen--Dice and categorical cross-entropy loss functions.
Data augmentation included random cropping to 320 pixel squares, random
orientation, scaling, perspective, contrast and brightness shifts and
image blurring. During final model evaluation we report the IoU of our
out-of-sample test datasets. The optimized model was used to classify
the complete orthomosaic using a moving window approach with a step size
of 110 pixels and a majority vote across overlapping areas to limit
segmentation edge effects. We refer to Figure 4 for a synoptic overview
of the full deep learning learning workflow. In addition, we used the
first acquisition of a recent pan-chromatic Geo-Eye 1 stereo pair
(GeoEye, 2012, Thornton, Colorado, U.S.A., order 737537, 2011-11-11
8:55h GMT or 9:55h local time) to classify and visually assess the
robustness of the algorithm on contemporary remote sensing imagery with
the same moving window approach.

\begin{figure}
\includegraphics[width=1\linewidth]{./figures/synthetic_landscape} \caption{An example synthetic landscape, combining homogeneous forest and non-forest images into a patchy landscape using random gaussian field based masks. The left panel shows a combined synthetic landscape, while the right panel shows the corresponding forest (black) and non-forest (white) labels.}\label{fig:unnamed-chunk-3}
\end{figure}

>>>>

%=================================================================
\documentclass[remote sensing,article,submit,moreauthors,pdftex]{mdpi}

% If you would like to post an early version of this manuscript as a preprint, you may use preprint as the journal and change 'submit' to 'accept'. The document class line would be, e.g., \documentclass[preprints,article,accept,moreauthors,pdftex]{mdpi}. This is especially recommended for submission to arXiv, where line numbers should be removed before posting. For preprints.org, the editorial staff will make this change immediately prior to posting.

@@

%--------------------
% Class Options:
%--------------------
%----------
% journal
%----------
% Choose between the following MDPI journals:
% acoustics, actuators, addictions, admsci, aerospace, agriculture, agriengineering, agronomy, algorithms, animals, antibiotics, antibodies, antioxidants, applsci, arts, asc, asi, atmosphere, atoms, axioms, batteries, bdcc, behavsci , beverages, bioengineering, biology, biomedicines, biomimetics, biomolecules, biosensors, brainsci , buildings, cancers, carbon , catalysts, cells, ceramics, challenges, chemengineering, chemistry, chemosensors, children, cleantechnol, climate, clockssleep, cmd, coatings, colloids, computation, computers, condensedmatter, cosmetics, cryptography, crystals, dairy, data, dentistry, designs , diagnostics, diseases, diversity, drones, econometrics, economies, education, electrochem, electronics, energies, entropy, environments, epigenomes, est, fermentation, fibers, fire, fishes, fluids, foods, forecasting, forests, fractalfract, futureinternet, futurephys, galaxies, games, gastrointestdisord, gels, genealogy, genes, geohazards, geosciences, geriatrics, hazardousmatters, healthcare, heritage, highthroughput, horticulturae, humanities, hydrology, ijerph, ijfs, ijgi, ijms, ijns, ijtpp, informatics, information, infrastructures, inorganics, insects, instruments, inventions, iot, j, jcdd, jcm, jcp, jcs, jdb, jfb, jfmk, jimaging, jintelligence, jlpea, jmmp, jmse, jnt, jof, joitmc, jpm, jrfm, jsan, land, languages, laws, life, literature, logistics, lubricants, machines, magnetochemistry, make, marinedrugs, materials, mathematics, mca, medicina, medicines, medsci, membranes, metabolites, metals, microarrays, micromachines, microorganisms, minerals, modelling, molbank, molecules, mps, mti, nanomaterials, ncrna, neuroglia, nitrogen, notspecified, nutrients, ohbm, particles, pathogens, pharmaceuticals, pharmaceutics, pharmacy, philosophies, photonics, physics, plants, plasma, polymers, polysaccharides, preprints , proceedings, processes, proteomes, psych, publications, quantumrep, quaternary, qubs, reactions, recycling, religions, remotesensing, reports, resources, risks, robotics, safety, sci, scipharm, sensors, separations, sexes, signals, sinusitis, smartcities, sna, societies, socsci, soilsystems, sports, standards, stats, surfaces, surgeries, sustainability, symmetry, systems, technologies, test, toxics, toxins, tropicalmed, universe, urbansci, vaccines, vehicles, vetsci, vibration, viruses, vision, water, wem, wevj

%---------
% article
%---------
% The default type of manuscript is "article", but can be replaced by:
% abstract, addendum, article, benchmark, book, bookreview, briefreport, casereport, changes, comment, commentary, communication, conceptpaper, conferenceproceedings, correction, conferencereport, expressionofconcern, extendedabstract, meetingreport, creative, datadescriptor, discussion, editorial, essay, erratum, hypothesis, interestingimages, letter, meetingreport, newbookreceived, obituary, opinion, projectreport, reply, retraction, review, perspective, protocol, shortnote, supfile, technicalnote, viewpoint
% supfile = supplementary materials

%----------
% submit
%----------
% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted. This will only make changes to the frontpage (e.g., the logo of the journal will get visible), the headings, and the copyright information. Also, line numbering will be removed. Journal info and pagination for accepted papers will also be assigned by the Editorial Office.

%------------------
% moreauthors
%------------------
% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.

%---------
% pdftex
%---------
% The option pdftex is for use with pdfLaTeX. If eps figures are used, remove the option pdftex and use LaTeX and dvi2pdf.

%=================================================================
\firstpage{1}
\makeatletter
\setcounter{page}{\@firstpage}
\makeatother
\pubvolume{xx}
\issuenum{1}
\articlenumber{5}
\pubyear{2019}
\copyrightyear{2019}
%\externaleditor{Academic Editor: name}
\history{Received: date; Accepted: date; Published: date}
\updates{yes} % If there is an update available, un-comment this line

%% MDPI internal command: uncomment if new journal that already uses continuous page numbers
%\continuouspages{yes}

@@

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{Land Use and Land Cover change (LULCC) of African rainforest contribute
to global carbon emissions. Yet, most historical estimates of LULCC and
their carbon emissions rely on non-spatially explicity data in the
pre-satellite era (\textless{} 1972). Past studies have expanded this
pre-satellite time frame with declassified satellite surveillance data
or aerial surveys, but none cover the Congo Basin. Here, we use
historical aerial survey photos to map the extent and structure of LULCC
around Yangambi, central Congo Basin, in 1958. Our study leveraged
Structure-from-Motion to generate a large orthomosaic covering 82800 ha,
counting 733 million pixels, at a sub meter resolution and
geo-referenced to \textasciitilde{}4.7 \(\pm\) 4.3 m. Primary forest in
the orthomosaic was classified with a Convolutional Neural Network Deep
Learning approach. We used these data to quantify LULCC, landscape and
canopy complexity and scale above ground biomass between historical and
contemporary condition. We show a shift from historical highly
structured industrial deforestation in 1958 (16200 ha), to contemporary
smallholder farming clearing (8800 ha), increasing landscape
fragmentation and forest edges. Efforts to quantify canopy texture and
their link to carbon storage had limited to no success. Our analysis
provides an insight in key spatial and temporal patterns of
deforestation and reforestation at a multi-decadal scale and provide a
historical context to land-cover and land-use change spatial patterns
for past and ongoing field research in the area.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Diversity
%\LSID{\url{http://}}

@@

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\usepackage{wrapfig} \usepackage{graphics} \usepackage{setspace} \doublespacing \usepackage{gensymb} \usepackage{booktabs}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Tropical ecosystem services are severely impacted by deforestation and
forest degradation \citep{hansen2013, houghton2012, tyukavina2015}. Not
only does tropical forest Land-Use and Land-Cover Change (LULCC)
constitute 10 to 15\% of the total global emissions
\citep{vanderwerf2009}, changes in forest fragmentation affect the
forest structure and function. Strong fragmentation effects decreases
the number of large trees along forest edges
\citep{laurance2000, magnago2015}, while species composition and
biodiversity are equally negatively affected
\citep{poorter2006, barlow2016, vandeperre2018}. Estimates show that
31\% of carbon emissions are caused by edge effects alone
\citep{brinck2017}.

As such, anthropogenic factors through LULCC heavily affect forest
structure and functioning \citep{fauset2017, brinck2017, didham1999}.
Accurate estimates of LULCC and forest canopy structure are therefore
imperative to estimate carbon emissions and other ecosystem services
\citep{houghton2012, hansen2013}. Remote sensing products have been key
input as they provide accurate spatial information to help estimate
carbon emissions \citep{mitchard2018, hansen2013}. High resolution
aerial images provide scientists tools to monitor forest extent,
structure and carbon emissions as canopy texture is linked to above
ground biomass \citep{Ploton2012, Couteron2005, Barbier2010}. Yet, these
estimates are often limited in time to recent decades
\citep{defries2002, houghton2012, hansen2013, achard2014}.

Historical estimates \textbf{of Land-Use and Land-Cover (LULC)}, in the
pre-satellite era (\textless{} 1972), exist but generally rely on
non-spatially explicit data (i.e socio-economic data)
\citep{houghton2012, ramankutty1999, defries2002, sader1988}. Efforts
have been made to use other geo-spatial data sources such as historical
maps \citep{willcock2016}, declassified CORONA satellite surveillance
data across the US and central Brazil \citep{song2015} as well as aerial
surveys in post world war II German \citep{nita2018}. Survey data across
the African continent is less common, inaccessible or both. Some studies
do exist, as \citet{buitenwerf2012}, \citet{hudak1998} and
\citet{okeke2006} used aerial survey images to map vegetation changes in
South African savannas and the Adulam Nature Reserve, respectively,
whilst \citet{frankl2015} and \citet{nyssen2016} mapped the Ethiopian
highlands of 1930's.

Despite the existence of large historical archives of aerial survey
imagery for the central Congo Basin (Figure 1, Appendix Figure 3) as of
yet no studies have valorized these data. The lack of a consistent
valorization effort is unfortunate as the African rainforest is the
second largest on Earth and covers \textasciitilde{}630 million ha,
representing up to 66 Pg of carbon storage \citep{Lewis2009b}, and
currently loses forest at an increasing pace \citep{butsic2015}. Given
the impact of LULCC on the structure and functioning of central African
tropical forests, and their influence on both carbon dynamics
\citep{Kearsley2013} and biodiversity \citep{vandeperre2018}, accurate
long term reporting of historical forest cover warrants more attention
\citep{willcock2016}.

Here, we use a combination of historical aerial photography and
contemporary remote sensing data to map long-term changes in the extent
and structure of the tropical forest surrounding Yangambi, in the
central Congo Basin, effectively linking the start of the antrophocene
\citep{lewis2015} with current assessments. Yangambi was, and remains, a
focal center of forest and agricultural research and development in the
central Congo Basin. Past research in the region allows for thorough
assessment of land-use and land-cover change from a multi-disciplinary
point of view, confronting us with complex deforestation and land-use
patterns.

We leverage Structure-from-Motion to generate a large orthomosaic of
historical imagery and develop a Convolutional Neural Network based
forest cover mapping approach, based upon a semi-supervised generated
dataset extensively leveraging data augmentation. Our methodology aims
to provides a historical insight into important LULCC spatial patterns
in Yangambi, such as fragmentation and edge complexity. We further
contextualize the influence of changes in the forest's life history on
past and current research into Above Ground Carbon (AGC) storage
\citep{Kearsley2013} and biodiversity \citep{vandeperre2018} in the
central Congo Basin. Our fast scalable mapping approach for historical
aerial survey data, using limited supervised input, would further
support long-term land-use and land-cover change analysis across the
central Congo Basin.

@@

Data for the central Congo Basin region, surrounding Kisangani, were
collected in several flights during the dry season of 1958 and 1959
(from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959
respectively, see Appendix Figure 1) to generate topographic maps of the
area, supervised by the ``Institut Géographique du Congo Belge'' in
Kinshasa (then Léopoldville). Black-and-White infrared images (0.4 - 0.9
\(\mu\)m) were gathered along flight paths running mostly from west to
east, between \textbf{09:00 and 11:00} local time. Along a flight path
continuous images were taken using a Wild Heerburgg RC5a (currently
Leica Geosystems) with an Aviogon lens assembly (114.83mm / f 5.6, with
a 90\degree view angle) resulting in square photo negative of 180 by 180
mm. Flights were flown at an average absolute altitude of
\textasciitilde{}5200 m above sea level, covering roughly 18 530
km\textsuperscript{2} at an approximate scale of 1/40 000. The use of
the integrated autograph system ensured timely acquisition of pictures
with a precise overlap (\textasciitilde{}1/3) between images. This large
overlap between images together with flight parameters would allow
post-processing, using stereographs, to create accurate topographic
maps. Original data from this campaign are stored in the
\href{https://www.africamuseum.be/en}{Royal Museum for Central Africa}
in Tervuren, Belgium (Figure 1).

\begin{figure}

{\centering \includegraphics[width=0.75\linewidth]{./figures/print_examples}

}

\caption{A box of historical aerial photographs (left) and a single aerial photograph (right) showing part of the Congo river. Note the meta-data provided in the margins of the image such as acquisition time and fleight height and the solar glare on the Congo river.}\label{fig:unnamed-chunk-1}
\end{figure}

@@

\begin{figure}

{\centering \includegraphics[width=0.75\linewidth]{./figures/flight_paths}

}

\caption{Overview of the historical flight paths during aerial photo acquisition and ancillary data used in this study. The bounding box of the orthomosaic data presented in this study is shown as a rectangle (23x36 km). The outline of a recent high-resolution Geo-eye panchromatic image is shown as a dashed dark grey rectangle (10x10km). The location of various permanent sampling plots are shown as x, +, and open squares and triangles for the mixed, mono-dominant and edge plots respectively. The grey polygon delineates the current day Yangambi Man and Biosphere reserve. The inset, top right, situates the greater Yangabmi region (white rectangle) with the DR Congo. The full flight plan and details are shown in Appendix Figure 1 and 2}\label{fig:unnamed-chunk-2}
\end{figure}

@@

All selected images, covering the Yangambi area, were contact prints as
original negatives of the prints were not available. Images were scanned
at a resolution exceeding their original resolution (or grain) at the
maximal physical resolution of an Epson A3 flatbed scanner (i.e.~2400
dpi or 160 MP per image) and saved as lossless tiff images. Data were
normalized using contrast limited histogram equalization
\citep{zuiderveld1994} with a window size of 32 and a clip limit of 1.5.
Fiduciary marks were used to rectify and downsample the images into
square 7700x7700 pixel images (\textasciitilde{}1200 dpi, 81 MP). This
resulted in a dataset with digital images at a resolution that remained
above the visible grain of the photographs, whilst the reduced image
size facilitated easier file handling and processing speed.

@@

We calculated the orthomosaic using a low resolution point cloud and
digital elevation map (DEM). Additional ground control points were
provided to assist in the referencing of image and constrain the
optimization routine used in the SfM algorithm. Ground control points
consisted of permanent structures which could be verified in both old
and new aerial imagery (i.e.~ESRI World Imagery) \textbf{and consisted
of corner points of build structures (e.g.~a building, bridge or
swimming pool etc.).} Although clouds were removed during the SfM
routine we did not mask all clouds in the final orthomosaic to maximize
forest coverage. The final scene was cropped, to provide consistent wall
to wall coverage of the reconstructed scene. The orthomosaic was
exported as a geotiff for further georeferencing in QGIS
\citep{qgisdevelopmentteam2018} using the georeferencer plugin (version
3.1.9) and additional ESRI World Imagery high resolution reference data.
We used 3\textsuperscript{th} degree polynomial and 16 ground control
points to correct the final image. Ground control points, raw image data
and final processed image are provided in addition to measures of
uncertainty such as mean and median error across all ground control
points. All subsequent analysis are executed on the final geo-referenced
orthomosaic or subsets of it.

@@

\hypertarget{classifying-land-use-and-land-cover}{%
\subsubsection{Classifying Land-Use and
Land-Cover}\label{classifying-land-use-and-land-cover}}

\hypertarget{model-training}{%
\paragraph{\texorpdfstring{\emph{Model
training}}{Model training}}\label{model-training}}

We automatically delineated all natural forest in the historical data,
thus excluding tree plantations, thinned or deteriorated forest stands
which \textbf{showed visible canopy cover loss}, fields and buildings.
We used the Unet Convolutional Neural Net (CNN, \citet{ronneberger2015})
architecture implemented in Keras \citep{chollet2015} with an
efficientnetb3 pre-processing backbone \citep{yakubovskiy2019} running
on TensorFlow \citep{martinabadi2015} to train a binary classifier
(i.e.~forest or non-forested). Training data were collected from the
orthomosaic by randomly selecting 513 pixel square tiles from
\textbf{locations within} homogeneous forested or non-forested polygons
in the historical orthomosaic \textbf{(Figure 5). Separate polygons were
selected for training, testing and validation purposes. Validation
polygons were sampled 300 times, while both testing and validation
polygons were sampled at 100 random locations. Tiles extracted from
locations close to the polygon border at times contained mixed cover
types. Tiles with mixed cover types were removed from the list of source
tiles (Table 1).} Homogeneous \textbf{source} tiles were combined in
synthetic landscapes using a random gaussian field based binary mask
(Figure 3). We generated 5000 synthetic landscapes for training, while
500 landscapes were generated for both the validation and the testing
datasets for a total of 6000 synthetic landscapes. In order to limit
stitch line misclassifications, along the seams of mosaicked images, we
created synthetic landscapes with different forest tiles to mimick
forest texture transitions. We applied this technique to 10\% of the
generated synthetic landscapes \textbf{(across training, validation and
testing data)}.

\caption{\label{tab:unnamed-chunk-3}Number of source tiles used for the generation of synthetic landscapes.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
  & training & testing & validation\\
\midrule
forest & 300 & 100 & 100\\
non-forest & 294 & 92 & 84\\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{./figures/synthetic_landscape}

}

\caption{An example synthetic landscape, combining homogeneous forest and non-forest images into a patchy landscape using random gaussian field based masks. The left panel shows a combined synthetic landscape, while the right panel shows the corresponding forest (black) and non-forest (white) labels.}\label{fig:unnamed-chunk-4}
\end{figure}
